ShrinkNets: Learning Network Size while Training

### 2018-02-09 paper/ICML/appendix.tex

\begin{figure}[t]
\begin{center}
\includegraphics[width=.7\columnwidth]{regressions}
\vspace*{-5mm}
\caption{\label{sparsity_accuracy}Loss/Sparsity trade off comparison between Group Sparsity and Shrinknets on linear and logistic regression. From top to bottom and left to right we show the results for \texttt{scm1d}, \texttt{oes97}, \texttt{gina\_prior2} and \texttt{gsadd}.}

\end{center}
\vspace*{-4mm}
\end{figure}

@@

\end{proof}
\subsection{Multi-Target Linear and Multi-Class Logistic regressions experiments}
As we showed, Group sparsity share similarities with our method, and we claim
that ShrinkNets are a relaxation of Group sparsity.  In this experiment we want
to compare the two approaches.  We decided to focus on multi-target linear
regression because in the single target case, groups in the Group Sparsity
problem would have a size of one ($\bm{A}$ would be a vector in this case).

The evaluation will be done on two datasets \texttt{scm1d} and \texttt{oes97}
\cite{Spyromitros-Xioufis2016} for linear regressions and we will use \texttt{gina\_prior2} \cite{4371065} and
the \textit{Gas Sensor Array Drift Dataset} \cite{VERGARA2012320} (that we shorten in
\texttt{gsadd}) for logistic regressions.

For each dataset we fit with different regularization parameters and measure
the error and sparsity obtained after convergence. In this context we define
sparsity as the ratio of columns that have all their weights under $10^{-3}$ in
absolute value. Regularization parameters were chosen in order to obtain the
widest sparsity spectrum. Loss is normalized depending on the problem to be in
the $[0, 1]$ range. We summarized the results in \cref{sparsity_accuracy}. From
our experiments it is clear that ShrinkNets can fit the data closer than Group
Sparsity for the same amount of sparsity. The fact that we are able to reach
very low loss demonstrate that even if our objective function is non convex, in
practice it works as good or better as convex alternatives.

>>>>

%\begin{figure}[t]
%\begin{center}
%\includegraphics[width=.7\columnwidth]{regressions}
%\vspace*{-5mm}
%\caption{\label{sparsity_accuracy}Loss/Sparsity trade off comparison between Group Sparsity and Shrinknets on linear and logistic regression. From top to bottom and left to right we show the results for \texttt{scm1d}, \texttt{oes97}, \texttt{gina\_prior2} and \texttt{gsadd}.}
%
%\end{center}
%\vspace*{-4mm}
%\end{figure}

@@

\end{proof}

%\subsection{Multi-Target Linear and Multi-Class Logistic regressions experiments}
%As we showed, Group sparsity share similarities with our method, and we claim
%that ShrinkNets are a relaxation of Group sparsity.  In this experiment we want
%to compare the two approaches.  We decided to focus on multi-target linear
%regression because in the single target case, groups in the Group Sparsity
%problem would have a size of one ($\bm{A}$ would be a vector in this case).
%
%The evaluation will be done on two datasets \texttt{scm1d} and \texttt{oes97}
%\cite{Spyromitros-Xioufis2016} for linear regressions and we will use \texttt{gina\_prior2} \cite{4371065} and
%the \textit{Gas Sensor Array Drift Dataset} \cite{VERGARA2012320} (that we shorten in
%\texttt{gsadd}) for logistic regressions.
%
%For each dataset we fit with different regularization parameters and measure
%the error and sparsity obtained after convergence. In this context we define
%sparsity as the ratio of columns that have all their weights under $10^{-3}$ in
%absolute value. Regularization parameters were chosen in order to obtain the
%widest sparsity spectrum. Loss is normalized depending on the problem to be in
%the $[0, 1]$ range. We summarized the results in \cref{sparsity_accuracy}. From
%our experiments it is clear that ShrinkNets can fit the data closer than Group
%Sparsity for the same amount of sparsity. The fact that we are able to reach
%very low loss demonstrate that even if our objective function is non convex, in
%practice it works as good or better as convex alternatives.

### 2018-02-09 paper/ICML/approach.tex

In this section we describe the ShrinkNets approach, which allows us to learn
the network size as part of the training process. The key idea is the addition
of \emph{switch layers} added after each layer in the network;  these layers
allow us to disable certain neurons during training.  We start by introducing the
basic method, and then explain how to adapt the training procedure to support this new
layer.

@@

At a high-level, out approach consists of two interconnected stages.
The first stage identifies neurons that do not improve the
prediction accuracy of the network and deactivates them. The second stage then
removes neurons from the network, shrinking its
size and yielding faster inference times. Broadly, these stages work as follows:

\noindent\textbf{Deactivating Neurons On-The-Fly: }During the first stage,
ShrinkNets applies an on/off switch to every neuron of an initially over-sized
network. We model the on/off switches by multiplying each input (or output) of
each layer by a parameter $\bm{\beta}\in\{0,1\}$. A value of $0$
will deactivate the neuron, while $1$ will let the signal go through. These
switches are part of a new layer, called the \textbf{switch layer}; this layer applies 
to fully connected as well as convolutional layers.

Our objective is to minimize the number of ``on'' switches to reduce the model size as much
as possible while preserving prediction accuracy. This can be achieved by jointly
minimizing the objective of the network and a factor of the L0 norm of the
vector containing all the switches. Because finding an optimal binary
assignment is an NP-Hard problem, we relax the problem by allowing
 $\bm{\beta}$ to be a real number
instead of a binary value and constrain it using the L1 instead of L0 norm.

\noindent\textbf{Neuron Removal: } During this stage, the neurons that are
deactivated by the switch layers are actually removed from the network,
effectively shrinking the network size. This step improves inference times, as
we demonstrate in our evaluation. We choose to remove neurons at training time
because we have observed that this allows the remaining active neurons to adapt
to the new network architecture. Existing techniques focus on neuron removal
after training, and require an extra fine-tuning process to compensate for the
removal.\srm{cite something?} \gl{I think we are fine or we could actually
cite any of these paper}

In the remainder of this section we describe in detail the switch layer as well
as and the training process for ShrinkNets, and then describe the removal process in Section~\ref{neuron_killing}.

@@

We want to tune the size of $L$ by applying a switch layer, $S$, containing $c$
switches.  The switch
layer is parametrized by a vector $\beta \in \mathbb{R}^c$ such that the result
of applying $S$ to $L(\bm{x})$ is a also a tensor size $\left(c \times d_1
\times \dots \times d_n\right)$ such that: 
%Suppose we wish to tune the size of $L$ by applying a switch layer.
%A switch layer $S$ applied to the output of $L$ can be parametrized by a 
%vector 
\begin{equation} 
S_{\beta}(L(\bm{x}))_{i,...} = \bm{\beta}_iL(\bm{x})_{i, ...} \forall i \in [1\ldots c]
\end{equation}
Once passed through the switch layer, each output channel $i$
produced by $L$ is scaled by the corresponding $\bm{\beta}_i$. Note that when
$\bm{\beta}_i = 0$, the $i^{\text{th}}$ channel is multiplied by zero and will not
contribute to any computation after the switch layer. If this happens, we say
the switch layer has {\it deactivated} the neuron corresponding to channel $i$ of layer $L$.

We place switch layers after each layer whose size we wish to tune; these are
typically fully connected and convolutional layers. We discuss next how to train
ShrinkNets.

\subsection{Training ShrinkNets}

@@

%Formally, 
For training, we need to account for the effect of the switch layers on the loss
function. The effect of switch layers can be expressed in terms of a sparsity constraint
that pushes values in the $\bm{\beta}$ vector to 0. In this way, given a neural network
parameterized by weights $\bm{\theta}$ and switch layer parameters $\bm{\theta}$, we
optimize ShrinkNet loss as follows 
\begin{equation}
\small
  L_{SN}(\bm{x},\bm{y};\bm{\theta}, \bm{\beta}) = L(\bm{x}, \bm{y}; \bm{\beta}) +
  \lambda\norm{\beta}_1 + \lambda_2\norm{\bm{\theta}}_p^p
\end{equation}
This expression augments the regular training loss with a regularization
term for the switch parameters and another on the network weights.

We next analyze why such loss function works to train ShrinkNets and its
connection to group sparsity:

>>>>

In this section we describe the \shrink approach, which allows us to learn
the network size as part of the training process. The key idea is the addition
of \swls after each layer in the network;  these layers
allow us to selectively deactivate neurons during training.  We start by introducing the
basic method, and then explain how to adapt the training procedure to support this new
layer.

@@

At a high-level, our approach consists of two interconnected stages.
The first stage identifies neurons that do not improve the
prediction accuracy of the network and deactivates them. 
The second stage then {\it removes} neurons from the network (explicitly
shrinking weight matrices and updating optimizer state) thus leading to 
smaller networks and also faster inference times.

\noindent\textbf{Deactivating Neurons On-The-Fly: }During the first stage,
\shrink applies an on/off switch to every neuron of an initially over-sized
network. We model the on/off switches by multiplying each input (or output) of
each layer by a parameter $\bm{\beta}\in\{0,1\}$. A value of $0$
will deactivate the neuron, while $1$ will let the signal go through. These
switches are part of a new layer, called the \swl; this layer applies 
to fully connected as well as convolutional layers.

Our objective is to minimize the number of ``on'' switches to reduce the model size as much
as possible while preserving prediction accuracy. This can be achieved by jointly
minimizing the training loss of the network and applying an L0 norm to the 
$\bm{\beta}$ parameters of the \swl. 
Since minimizing the L0 norm is an NP-Hard problem, we instead relax the constraint
to an L1 norm by constraining $\bm{\beta}$ to be a real number
instead of a binary value.
% Because finding an optimal binary
% assignment is an NP-Hard problem, we relax the problem by allowing
%  $\bm{\beta}$ to be a real number
% instead of a binary value and constrain it using the L1 instead of L0 norm.

\noindent\textbf{Neuron Removal: } During this stage, the neurons that are
deactivated by the switch layers are actually removed from the network,
effectively shrinking the network size. This step improves inference times, as
we demonstrate in our evaluation. We choose to remove neurons at training time
because we have observed that this allows the remaining active neurons to adapt
to the new network architecture and we can avoid a post-training step to prune
deactivated neurons.
% Existing techniques focus on neuron removal
% after training, and require an extra fine-tuning process to compensate for the
% removal.
% \srm{cite something?} \gl{I think we are fine or we could actually
% cite any of these paper}

In the remainder of this section we describe in detail the switch layer as well
as and the training process for \shrink, and 
then describe the removal process in Section~\ref{neuron_killing}.

@@

We want to tune the size of $L$ by applying a \swl, $S$, containing $c$
switches.  The \swl is parametrized by a vector $\bm{\beta} \in \mathbb{R}^c$ such that the result
of applying $S$ to $L(\bm{x})$ is a also a tensor size $\left(c \times d_1
\times \dots \times d_n\right)$ such that: 
%Suppose we wish to tune the size of $L$ by applying a switch layer.
%A switch layer $S$ applied to the output of $L$ can be parametrized by a 
%vector 
\begin{equation} 
S_{\beta}(L(\bm{x}))_{i,...} = \bm{\beta}_iL(\bm{x})_{i, ...} \forall i \in [1\ldots c]
\end{equation}
Once passed through the switch layer, each output channel $i$
produced by $L$ is scaled by the corresponding $\bm{\beta}_i$. Note that when
$\bm{\beta}_i = 0$, the $i^{\text{th}}$ channel is multiplied by zero and will not
contribute to any computation after the switch layer. If this happens, we say
the switch layer has {\it deactivated} the neuron corresponding to channel $i$ of layer $L$.

We place \swl after each layer whose size we wish to tune; these are
typically fully connected and convolutional layers. We discuss next how to train
\shrink.

\subsection{Training \shrink}

@@

%Formally, 
For training, we need to account for the effect of the \swls on the loss
function. The effect of \swls can be expressed in terms of a sparsity constraint
that pushes values in the $\bm{\beta}$ vector to 0. In this way, given a neural network
parameterized by weights $\bm{\theta}$ and switch layer parameters $\bm{\theta}$, we
optimize \shrink loss as follows 
\begin{equation}
\small
  L_{SN}(\bm{x},\bm{y};\bm{\theta}, \bm{\beta}) = L(\bm{x}, \bm{y}; \bm{\beta}) +
  \lambda\norm{\beta}_1 + \lambda_2\norm{\bm{\theta}}_p^p
\end{equation}
This expression augments the regular training loss with a regularization
term for the switch parameters and another on the network weights.

We next analyze why such loss function works to train \shrink and its
connection to group sparsity:

### 2018-02-09 paper/ICML/evaluation.tex

%\srm{Summarize more quantitatively -- best model performance, model performance
%drop off for a factor of 10 reduction in seize, etc. so that we can repeat a
%quantitative claim in the intro, i.e., ``Here, the best performing ShrinkNets
%models has 9x.x\% accuracy while the best static model has 9x.x\% accuracy,
%while the ShrinkNets model is x times smaller.  Further, note that ShrinkNets
%finds a model with 9x.x\% accuracy that is y times than the best performing
%model.  Also, as on CIFAR10, if we give up just 1\% error, ShrinkNets finds a
%model that is z times smaller.'' (If we do this, would be good to highlight
%these points in the figure somehow)}

\subsubsection{Summary}

We  demonstrated that it is possible to achieve networks with good accuracy
when using ShrinkNets both when the network space cannot be explored entirely
(\texttt{CIFAR10}) and when it can, e.g., \texttt{COVERTYPE}. The most important
result is not that ShrinkNets finds networks of good accuracy, but that those
networks are much smaller than those found by a static method. The impact of the
network size on inference time is the subject of our next evaluation goal.

@@

\ra{the following is interesting, but how is it related to the above?}
ShrinkNets effectively explores the frontier of model size and accuracy. For a
given target accuracy, the size needed is significantly smaller than when we use the
"channel doubling" heuristic commonly used to size convolutional neural networks.
This suggests that this conventional heuristic may not in fact be optimal,
especially when looking for smaller models.  Empirically we observed this to
often be the case.  For example, during our experimentations on the
\texttt{MNIST} \cite{Lecun1998} and \texttt{FashionMNIST} \cite{Xiao2017}
datasets (not reported here due to space constraints), we observed that even
though these datasets have the same number of classes, input features, and
output distributions, for a fixed $\lambda$ \textit{ShrinkNets} converged to
considerably bigger networks in the case of \texttt{FashionMNIST}. This evidence
shows that optimal architecture not only depends on the output distribution or
shape of the data but actually reflects the dataset.  This makes sense, as
\texttt{MNIST} is a much easier problem than \texttt{FashionMNIST}.

>>>>

\noindent\textbf{Summary.} We demonstrated that it is possible to achieve
networks with good accuracy when using ShrinkNets both when the network space
cannot be explored entirely (\texttt{CIFAR10}) as well as when it can, e.g.,
\texttt{COVERTYPE}. The most important result is not that ShrinkNets finds
networks of good accuracy, but that those networks are much smaller than those
found by a static method. The impact of the network size on inference time is
the subject of our next evaluation goal.

@@

ShrinkNets effectively explores the frontier of model size and accuracy. For a
given target accuracy, the size needed is significantly smaller than when we use the
"channel doubling" heuristic commonly used to size convolutional neural networks.
This suggests that this conventional heuristic may not in fact be optimal,
especially when looking for smaller models.  Empirically we observed this to
often be the case.  For example, during our experimentations on the
\texttt{MNIST} \cite{Lecun1998} and \texttt{FashionMNIST} \cite{Xiao2017}
datasets (not reported here due to space constraints), we observed that even
though these datasets have the same number of classes, input features, and
output distributions, for a fixed $\lambda$ \textit{ShrinkNets} converged to
considerably bigger networks in the case of \texttt{FashionMNIST}. This evidence
shows that optimal architecture not only depends on the output distribution or
shape of the data but actually reflects the dataset.  This makes sense, as
\texttt{MNIST} is a much easier problem than \texttt{FashionMNIST}.

### 2018-02-09 paper/ICML/introduction.tex

Of course, the problem of finding compressed or small networks is not new.
Existing techniques typically aim make a pre-trained neural network smaller
by taking one of two approaches:  either by
applying quantization \cite{quant} or code compilation~\cite{ma2016compilation}
techniques that can be applied blindly to any network, or by analyzing the
structure of the network and systematically pruning
connections or neurons~\cite{han2015deepcompression,Cun}
based on some loss function.  Although these techniques can substantially
reduce model size, they have several drawbacks.  First, they often negatively
impact model quality.  Second, they can (surprisingly) negatively impact
inference time as they transform dense matrix operations into sparse ones,
which can be substantially slower to execute on modern hardware
\cite{han2015deepcompression}, especially GPUs which do not efficiently support sparse
linear algebra.  Third, these techniques generally start by optimizing a
particular architecture for prediction performance, and then, as a post-
processing step, applying compression  to generate a smaller model that meets the
resource constraints of the deployment setting.  Because the network
architecture is essentially fixed during this post-processing,   model architectures
that work better in small settings may be missed -- this is especially true in
large networks like many-layered CNNs,  where it is infeasible to try explore
even a small fraction of possible network configurations.

@@

\item We implement our technique so as to remove entire neurons, therefore
leading not only to small and well-performing networks, but also dense, which
means higher inference performance.

>>>>

Of course, the problem of finding compressed or small networks is not new.
Existing techniques to make a pre-trained neural network smaller can be categorized in two approaches: (1) quantization \cite{quant} and code compilation~\cite{ma2016compilation}, 
techniques that can be applied blindly to any network, and (2) techniches which analyze the
structure of the network and systematically prune
connections or neurons~\cite{han2015deepcompression,Cun}
based on some loss function. While the first category is always useful, but only has limited 
impact on the network size, the second category can reduce the model size much more but has several 
drawbacks:   First, those techniques often negatively
impact model quality.  Second, they can (surprisingly) negatively impact
inference time as they transform dense matrix operations into sparse ones,
which can be substantially slower to execute on modern hardware
\cite{han2015deepcompression}, especially GPUs which do not efficiently support sparse
linear algebra.  Third, these techniques generally start by optimizing a
particular architecture for prediction performance, and then, as a post-
processing step, applying compression  to generate a smaller model that meets the
resource constraints of the deployment setting.  Because the network
architecture is essentially fixed during this post-processing,   model architectures
that work better in small settings may be missed -- this is especially true in
large networks like many-layered CNNs,  where it is infeasible to try explore
even a small fraction of possible network configurations.

@@

\item We implement our technique so as to remove entire neurons, therefore
leading not only to small and well-performing networks, but also dense, which
means higher inference performance. Furthermore, our switching layers used during training can be 
savely removed before the model is used for inference, leading to zero additional overhead.

### 2018-02-09 paper/ICML/paper.tex

% Use the following line for the initial blind version submitted for review:
\usepackage{icml2018}
\newcommand{\srm}[1]{\textcolor{red}{{\bf Sam:} #1}}
\newcommand{\ra}[1]{\textcolor{blue}{{\bf ra:} #1}}
\newcommand{\gl}[1]{\textcolor{violet}{{\bf Gl:} #1}}
\newcommand{\mpv}[1]{\textcolor{green}{{\bf MV:} #1}}
\newcommand{\shrink}{{\bf ShrinkNets}\xspace}
\newcommand{\swl}{{\bf SwitchLayer}\xspace}
\newcommand{\swls}{{\bf SwitchLayers}\xspace}
% If accepted, instead use the following line for the camera-ready submission:
%\usepackage[accepted]{icml2018}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Learning Neural Network size with ShrinkNets}

@@

\twocolumn[
\icmltitle{Learning Network Size while Training with ShrinkNets}

@@

\begin{abstract}
As neural networks are being increasingly deployed in different applications and
on different hardware, it has become increasingly important to optimize
inference time and model size along with model accuracy.  Most current work
optimizes model size, model accuracy and inference time in different stages and
results in suboptimal results at the cost of computational inefficiency.  In
this work, we propose a new techique called ShrinkNets that optimizes all three
of these metrics simultaneously by learning optimal network size at training
time.  Specifically our technique presents a new method to simultaneously
optimize network size and model performance by neuron-level pruning.
Neuron-level pruning not only produces much smaller networks for a given
accuracy threshold but also produces dense weight matrices that are efficient at
inference time.  By applying our technique to a variety of models with both
convolutional and fully connected layers, we show that ShrinkNets can reduce by
35\% the network size with respect to a static network for \texttt{CIFAR10}, and
it can achieve speedups during inference time of up to 40X. In addition, the
most accurate networks found by ShrinkNets are similar or more accurate than
those found by statically trained networks.  
\end{abstract}

>>>>

% Use the following line for the initial blind version submitted for review:
\usepackage{icml2018}
\newcommand{\srm}[1]{\textcolor{red}{{\bf Sam:} #1}}
\newcommand{\tim}[1]{\textcolor{red}{{\bf Tim:} #1}}
\newcommand{\ra}[1]{\textcolor{blue}{{\bf ra:} #1}}
\newcommand{\gl}[1]{\textcolor{violet}{{\bf Gl:} #1}}
\newcommand{\mpv}[1]{\textcolor{green}{{\bf MV:} #1}}
\newcommand{\shrink}{{\bf ShrinkNets}\xspace}
\newcommand{\swl}{{\bf SwitchLayer}\xspace}
\newcommand{\swls}{{\bf SwitchLayers}\xspace}
% If accepted, instead use the following line for the camera-ready submission:
%\usepackage[accepted]{icml2018}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{ShrinkNets: Learning Network Size while Training}

@@

\twocolumn[
\icmltitle{ShrinkNets: Learning Network Size while Training}

@@

\begin{abstract}
As neural networks are being widely in different applications and
on different hardware, it has become increasingly important to optimize
inference time and model size along with model accuracy.  Most current work
optimizes model size, model accuracy and inference time in different stages, 
resulting in suboptimal results and computational inefficiency.  In
this work, we propose a new technique called ShrinkNets that optimizes all three
of these metrics simultaneously by learning optimal network size at training
time.  Specifically our technique presents a new method to simultaneously
optimize network size and model performance by neuron-level pruning.
Neuron-level pruning not only produces much smaller networks for a given
accuracy threshold but also produces dense weight matrices that are efficient at
inference time.  By applying our technique to a variety of models with both
convolutional and fully connected layers, we show that ShrinkNets can reduce network size
by 35x in comparison to to a static network on \texttt{CIFAR10}, with
6x improvement in inference time. In addition, the
most accurate networks found by ShrinkNets are similar or more accurate than
those found by statically trained networks.  
\end{abstract}
